---
title: "CulexT DE"
format: html
---
# Setup

## Dependencies

1. Change directory to repo root in order for relative paths to be shorter.
2. Bioconductor packages must be checked explicitly for installation.
3. RStudio should detect standard packages being loaded via `library()`, and give you the option to install.  


```{r setup, message=F, warning=F}
knitr::opts_knit$set(root.dir = normalizePath("..")) # assuming wd is this file's directory: 04_DEseq2
library(BiocManager)
Biocpackages = c("DESeq2")
# check for installation, of packages
for (pkg in Biocpackages) {
    if (!require(pkg,character.only = T, quietly=T)) {
        BiocManager::install(pkg)
        library(pkg, character.only = T, quietly = T)
    }
}

# pipes, data and string manipulation
library(tidyverse, quietly=T)
library(dplyr, quietly=T)
library(stringr, quietly=T)
library(magrittr, quietly=T)
```

## Data

1. Function to handle column names in Hunter's data files.
2. Read Raw-counts to process data. Read Normalized counts for comparison.
```{r cleanUpNames function}
cleanUpNames <- function(colnames) {
  colnames %>%
    str_replace("^X", "T") %>%                  # Replace leading "X" → "T"
    str_remove("_S\\d+HSCounts\\.txt$")         # Remove suffix "_S5HSCounts.txt"
  
}
```

```{r read-csvs}
countfile = "Hunter_Results/EG_Raw_Counts.csv"
df = read.table(countfile, sep=',', header=T,row.names=1)
colnames(df) <- cleanUpNames(colnames(df))
count_mat <- df %>% as.matrix()
storage.mode(count_mat) <- "integer"

hunter_norm_counts = read.table("Hunter_Results/EG_Normalized_Counts.csv", sep = ',', header=T, row.names=1)
colnames(hunter_norm_counts) <- cleanUpNames(colnames(hunter_norm_counts))
```

### Build metadata table with factors

1. Parse Temperature (22,26,30), Infection (mock or WNV), and replicate (A,B,C) from column names.
2. Make variables into factors
3. Make `condition` as the combined factor levels of temperature and infection

```{r build-sample-metadata}
coldata <- tibble(sample = colnames(count_mat)) %>%
  mutate( # extract the factor values from the column names
    temperature = str_extract(sample, "(?<=T)\\d+"),     # 22, 26, 30
    infection   = str_extract(sample, "mock|WNV"),       # infection type
    replicate   = str_extract(sample, "[A-Z]$")          # replicate ID (A, B, C)
  ) %>%
  mutate( # make factors out of the levels
    temperature = factor(temperature, levels = c("26", "22", "30")), 
    infection = factor(infection, levels = c("mock", "WNV"))
  ) %>%
  mutate( # make the combined factor variable
    condition = factor(paste0("T", temperature, "_", infection))
  ) %>%
  column_to_rownames("sample")

coldata
```

# Analysis

1. Pass count matrix, column data, and formula into DESeq2. This does the basic model fitting.
2. Get the normalized counts to compare with Hunter's.

## DESeq2 object
```{r create-object}
dds <- DESeqDataSetFromMatrix(
  countData = count_mat,
  colData   = coldata,
  design    = ~ condition
)
dds <- DESeq(dds)

resultsNames(dds)
```

```{r norm counts}
dds <- estimateSizeFactors(dds)
norm_counts <- counts(dds, normalized = TRUE)
log_norm_counts <- log2(norm_counts + 1)
```

```{r compare-w-hunter}
max(abs(norm_counts - hunter_norm_counts)) # extremely small value
```

## Create a result object with all-pairwise comparisons between conditions

Given the factor levels of the combined variables (named "condition"), compute all the
pairwise contrasts of those values.

```{r all-pairwise}
# Get all levels of the condition factor
conds <- levels(colData(dds)$condition)

# Generate all unique pairwise combinations
comparisons <- combn(conds, 2, simplify = FALSE)

# For each pairwise contrast, extract results
res_list <- lapply(comparisons, function(x) {
  contrast_name <- paste(x[1], "vs", x[2])
  res <- results(dds, contrast = c("condition", x[1], x[2]))
  res_df <- as.data.frame(res) %>%
    rownames_to_column("gene_id") %>%
    mutate(comparison = contrast_name,
           compare_from = x[1], 
        compare_to = x[2])
  return(res_df)
})

# Combine all results into one tidy dataframe
all_res <- bind_rows(res_list)
```

### How many genes are significant for each condition/comparison

A p-adjust threshold of 0.1 was used uniformly on this dataset.

```{r global results}
# tabulate number of "significant" genes
all_res %>%
  filter(is.finite(padj) & padj < 0.1) %>%
  count(comparison, sort = TRUE)
```

```{r gene lists}
sig_rows <- all_res %>% filter(is.finite(padj) & padj < .1)
sig_table <- pivot_wider(sig_rows, id_cols=gene_id, names_from = comparison, values_from = padj) %>% column_to_rownames('gene_id')

all_table_padj <- pivot_wider(all_res, id_cols=gene_id, names_from = comparison, values_from = padj) %>% column_to_rownames('gene_id')
all_table_l2FC <- pivot_wider(all_res, id_cols=gene_id, names_from = comparison, values_from = log2FoldChange) %>% column_to_rownames('gene_id')

safe_min <- function(x) {
    if (any(is.finite(x))) {
        return(min(x, na.rm=T) < .1)
    } else {
        return(FALSE)
    }
}
which_sig_rows <- apply(all_table_padj, 1, safe_min) 
sig_table_padj <- all_table_padj[which_sig_rows,]
sig_table_l2FC <- all_table_l2FC[which_sig_rows,]
sig_table_mask <- sig_table_padj < .1
sig_table_mask[!is.finite(sig_table_mask)] <- FALSE
sig_table_l2FC[!sig_table_mask] <- NA

# save these files

all_res_trunc <- all_res %>% mutate( # long format file doesn't need so much
    baseMean = round(baseMean, digits = 9),
    log2FoldChange = round(log2FoldChange, digits = 9),
    lfcSE = round(lfcSE, digits = 9),
    stat = round(stat, digits = 9),
    pvalue = round(pvalue, digits = 9),
    padj = round(padj, digits = 9)
)
DATADIR= "04_DEseq2/DCK_results"
path = file.path(DATADIR, "EG_all_results_DCK.tsv")
all_res_trunc %>% write.table(path, sep="\t", row.names=F, quote=F)

path = file.path(DATADIR, "EG_all_results_table_DCK.tsv")
all_table %>% write.table(path, sep="\t", quote=F)

path = file.path(DATADIR, "EG_sig_padj_DCK.tsv")
sig_table_padj %>% write.table(path, sep="\t", quote=F)

path = file.path(DATADIR, "EG_sig_l2fc_DCK.tsv")
sig_table_l2FC %>% write.table(path, sep="\t", quote=F)
```

### Load and compare Hunter's output

Each of Hunter's comparisons are in a different file, so loop over them and compare
log2FoldChange, padj, and the number of significant genes with the ones computed above.

```{r compare results with Hunters}

comparison_df = data.frame()

root_dir = 'Hunter_Results/geneID_DifferentialExpressionResults'
for (Hfile in list.files(root_dir, pattern="*_FC.csv")) {
    filename_noext = str_remove(Hfile, "_DE_FC\\.csv$")
    # split on underscore
    parts = str_split(filename_noext, "_")
    condition_from = map_chr(parts, 1)
    condition_to   = map_chr(parts, 2)
    # convert to your target naming style (T22_mock, etc.)
    condition_from = str_replace(condition_from, "(\\D+)(\\d+)", "T\\2_\\1")
    condition_to   = str_replace(condition_to, "(\\D+)(\\d+)", "T\\2_\\1")
    # lowercase 'mock' only
    condition_from = str_replace(condition_from, "Mock", "mock")
    condition_to   = str_replace(condition_to, "Mock", "mock")

    Hdata = read.table(file.path(root_dir, Hfile), sep=',', header=T, row.names=1) %>%
        rownames_to_column("gene_id") %>% 
        arrange(gene_id)
    Ddata = all_res %>% filter(compare_from == condition_from & compare_to == condition_to) %>% 
            arrange(gene_id)

    largest_l2fc_diff = max(abs(Hdata$log2FoldChange - Ddata$log2FoldChange), na.rm=T)
    largest_padj_diff = max(abs(Hdata$padj - Ddata$padj), na.rm=T)
    n_Hunter_sig = sum(Hdata$padj < .1,na.rm=T)
    n_David_sig = sum(Ddata$padj < .1,na.rm=T)

    comparison_df <- rbind(
                    comparison_df,
                    data.frame(condition_from=condition_from,
                               condition_to=condition_to,
                               largest_l2fc_diff=largest_l2fc_diff,
                               largest_padj_diff=largest_padj_diff,
                               n_Hunter_sig=n_Hunter_sig,
                               n_David_sig=n_David_sig
                               )
    )

}

comparison_df
```